Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-12 manuscript/manuscript.tex
more theory section stuff

Our work is motivated by a well-known limitation of classic background
selection theory: it is only accurate for strongly deleterious mutations
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}. Given
statistical methods that fit genome-wide diversity to linked selection
processes rely on the classic BGS model to estimate the distribution of fitness
effects and rate of deleterious mutations, it is important to characterize and
remedy these shortcomings. However, extending background selection theory to
model weak negative selection is challenging, and it is worthwhile to develop
some intuition for how the classic background selection works and why this is
the case.

The classic background selection model \parencite{Charlesworth1993-gb}

First, the frequency trajectory of a weakly selected allele is
predominantly determined by stochastic evolutionary processes. When the changes
due to stochasticity are large, deterministic approximations that work well
under strong selection no longer capture selection dynamics. Second, weakly
selected alleles are impacted by two sources of evolutionary stochasticity:
genetic drift (i.e. non-heritable differences in offspring number and Mendelian
segregation) and genetic draft, the stochastic changes in allele frequency
caused by selection on the same genetic background
\parencite{Neher2013-dz,McVean2000-bt}.

Moreover, when the magnitude of stochastic changes can exceed the deterministic
changes caused by selection, mutations that decrease fitness now have a chance
of fixation \parencite{Kimura1962-su,Malecot1948-zv}. Because there is a steady
influx of weakly deleterious mutations, each with a non-zero chance of
substitution, there must be a corresponding substitution process. The rate of
this deleterious substitution process (e.g. ``Muller's ratchet") is

By contrast, when selection is assumed to be strong,
deterministic models of mutation and selection that ignore these sources of
stochasticity adequately capture the dynamics. Since weakly deleterious
mutations have some chance of fixation,

@@

describe how selection alters pairwise coalescence times rather than full
genealogy (since $\E[\pi] = 2\mu \E(T_2)$). Statistical methods that fit the
relationship between observed diversity to an expectation provided by a
particular theoretic model of the linked selection process.

Our work considers a well-known
shortcoming of the classic background selection model: that

The classic theoretic model of how negative or purifying selection changes the
mean coalescence time at linked neutral sites is the background selection model
\parencite{Charlesworth1993-gb}. In the simplest case without recombination,
this model assumes that deleterious mutations enter a segment.

@@

new deleterious mutations enter a region of $L$-basepairs at a rate of $\mu$
per basepair per generation with a selection coefficient $s$ in heterozygotes
that is sufficiently deleterious that it is destined to loss. Under this setup,
the number of segregating deleterious mutations in the region reaches a
stationary Poisson distribution with rate $\nicefrac{U}{s}$ ($U=\mu L$), which
is precisely the equilibrium number under a deterministic mutation-selection
balance models \parencite{Kimura1966-bk,Crow1970-wj}. Since any linear carrying
a deleterious mutation is destined to extinction, coalescence events can only
occur in the $N \exp(-\nicefrac{U}{s})$ individuals carrying zero deleterious
mutations. Working back through the ancestors of two lineages carrying $i$ and
$j$ deleterious mutations, both

>>>>

@@

Our work is motivated by a well-known limitation of classic background
selection theory: it is only accurate for strongly deleterious mutations
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}. Given
statistical methods that fit linked selection processes to genome-wide
diversity data rely on the classic BGS model to estimate the distribution of
fitness effects and rate of deleterious mutations, it is important to
characterize and remedy these shortcomings. However, extending background
selection theory to model weak negative selection is challenging, and it is
worthwhile to develop some intuition for why this is the case, how the classic
background selection works, and how the model of \textcite{Santiago2016-mu}
solves these problems and can be extended to genome-wide inference.

Both the classic background selection and SC16 models imagine a negative
selection process where a continual influx of deleterious mutations flow into
the population at a rate of $\mu$ per basepair per generation in a conserved
region of $L$ basepairs, such that the region-wide per generation mutation rate
is $U = \mu L$. Each mutation imposes an additive selective cost of $s$ in
heterozygotes and $2s$ in homozygotes. The classic background selection model
considers when mutations are sufficiently deleterious that they are destined to
extinction
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}. In
this case, the genealogy is well-approximated by a neutral coalescence process
with a rescaled effective population size of $N_0 = N \exp(-\nicefrac{U}{s})$.
This approximation works reasonably well for two reasons (though see
\textcite{Cvijovic2018-vd}). First, when selection is strong, the number of
deleterious mutations per haplotype has a stationary Poisson distribution with
mean equal to the equilibrium under the deterministic mutation-selection model
\parencite{Haldane1927-ga}. Since lineages carrying any strongly deleterious
mutations cannot contribute to the genealogy, coalescence events only occur in
the fraction $\exp(-\nicefrac{U}{s}$ free of mutations. Second, two present-day
lineages carrying any deleterious mutations trace their ancestry back to the
pool of $N_0$ mutation-free ancestors very quickly, since these harmful
mutations could only have been acquired recently by the lineage. On coalescence
timescales of $\sim 2N_0$, this time is negligible and can be ignored
(\textcite{Durrett2008-ql}, p. 213; \textcite{Good2014-yz}).

@@
